A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-27 paper.tex
tweaked the bounded section

At every round of the game, the \emph{environment} picks an
uncontrollable label, the \emph{controller} responds by choosing a controllable 
label and the game transitions into a new state according to $\delta$. 
A \emph{run} of a game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n,
c_n)$ is a chain of state and label pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$. 
A run is winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot E(x_i))$. 
In a \emph{bounded game} of rank $n$ all runs are restricted to length $n$, whereas unbounded 
games consider runs of infinite length. Since we consider only deterministic games, a run 
is uniquely described by a list of assignments to $L_c$ and $L_u$.

@@

\subsection{Abstract Game Trees}

@@

        \caption{AGT with strategy}

@@

                child {node [circle,draw] (sroot) {}
                    child {node [below left=0 and -10pt] (sa) {}
                        edge from parent [draw=none]
                    child {node [below right=0 and -10pt] (sb) {}
                        edge from parent [draw=none]
                \draw[unfixed] (sa.center) -- (sb.center);
                \draw[unfixed] (sroot.south) -- (sa.center);
                \draw[unfixed] (sroot.south) -- (sb.center);
        \caption{AGT with spoiling strategy}
        \label{fig:spoiling}

@@

We review the bounded synthesis algorithm by Narodytska et
al.~\cite{narodytska2014}, which is the main building block for our unbounded
algorithm.  The algorithm constructs a series of abstractions of the input
game.  An abstraction restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Together with a state-round pair
$\langle s,k\rangle$, an abstract game tree defines an \emph{abstract game}
played from this state.  Figure~\ref{fig:agt} shows an example abstract game
tree.  In the abstract game, the environment player is required to pick actions
from the tree, starting from the root node.  After reaching a leaf, it
continues playing unrestricted.  The tree in Figure~\ref{fig:agt} restricts the
initial environment action to the set $\{a, d\}$.  After choosing action $d$,
the environment reaches a leaf of the tree and continues playing unrestricted.
Alternatively, after choosing $a$, the environment is required to play action
$b$ in the next round.

Nodes of an abstract game tree are uniquely identified by the list of edge
labels along the path from the root to the node.  We identify an abstract game
tree with the set of its nodes.  For example, the tree in Figure~\ref{fig:agt}
can be written as $\{(),(d),(a),(a,b)\}$.  We denote $\textsc{leaves}(T)$ the
subset of leaf nodes of a tree $T$.

A \emph{partial strategy} $Strat: T \rightarrow L_c $ assigns a controllable
action to be played in each node of the abstract game tree.
Figure~\ref{fig:strategy} shows an example partial strategy.  The controller
starts by choosing action $\alpha$.  If the environment plays $a$, the
controller responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, \langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$p, \langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, \langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, \langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, \langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

Algorithm~\ref{alg:bounded} illustrates the bounded synthesis algorithm. The
algorithm takes a concrete game $G$ as an implicit argument. In addition, it
takes a player (controller or environment), state-round pair $\langle s,
k\rangle$ and an abstract game tree $T$ and returns a winning partial strategy
for it, if one exists.  The initial invocation of the algorithm takes the
initial state $\langle I, 0\rangle$ and an empty abstract game tree
$\emptyset$. Initially the solver is playing on behalf of the environment since
that player takes the first move in every step.  The empty game tree does not
constrain opponent moves, hence solving such an abstraction is equivalent to
solving the original concrete game.  The algorithm is organised as a
counterexample-guided abstraction refinement (CEGAR) loop.  The first step of
the algorithm uses the \textsc{findCandidate} function, described in detail
below, to come up with a candidate partial strategy for $T$. If it fails to
find a strategy, this means that no winning partial strategy exists for $T$.
If, on the other hand, a candidate partial strategy is found, we need to verify
if it is indeed winning for $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.  {Figure~\ref{fig:spoiling}} shows a spoiling
strategy discovered one of the leaves of the partial strategy.

>>>>

At every round of the game, the \emph{environment} picks an
uncontrollable label, the \emph{controller} responds by choosing a controllable 
label and the game transitions into a new state according to $\delta$. 
A \emph{run} of a game $(x_0, u_0, c_0), (x_1, u_1, c_1) \dots (x_n, u_n,
c_n)$ is a chain of state and label pairs s.t.\,$\delta(x_k, u_k, c_k, x_{k+1})$. 
A run is winning for the controller if $x_0 = I \land \forall i \in \{1..n\} (\lnot E(x_i))$. 
In a \emph{bounded game} with bound $n$ all runs are restricted to length $n$, whereas unbounded 
games consider runs of infinite length. Since we consider only deterministic games, a run 
is uniquely described by a list of assignments to $L_c$ and $L_u$.

@@

@@

        \caption{AGT with partial strategy}

@@

%    \begin{subfigure}[t]{.25\textwidth}
%        \centering
%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%            \node [circle,draw] (root){}
%                child {node [circle,draw] {}
%                    child {node [circle,draw] {}
%                        node [left=4pt] {$\epsilon$}
%                        edge from parent [fixed] node [left] {$b$}
%                    }
%                    node [left=4pt] {$\beta$}
%                    edge from parent [fixed] node [left] {$a$}
%                }
%                child {node [circle,draw] (sroot) {}
%                    child {node [below left=0 and -10pt] (sa) {}
%                        edge from parent [draw=none]
%                    }
%                    child {node [below right=0 and -10pt] (sb) {}
%                        edge from parent [draw=none]
%                    }
%                    node [left=4pt] {$\gamma$}
%                    edge from parent [fixed] node [right] {$d$}
%                }
%                node [right=4pt] {$\alpha$}
%                node [above=4pt] {$\langle s, k \rangle$};
%                \draw[unfixed] (sa.center) -- (sb.center);
%                \draw[unfixed] (sroot.south) -- (sa.center);
%                \draw[unfixed] (sroot.south) -- (sb.center);
%        \end{tikzpicture}
%        \captionsetup{width=.8\textwidth}
%        \caption{AGT with spoiling strategy}
%        \label{fig:spoiling}
%    \end{subfigure}%

@@

We review the bounded synthesis algorithm by Narodytska et
al.~\cite{narodytska2014}, which is the main building block of our unbounded
algorithm.  The algorithm constructs a series of abstractions of the input
game.  
An abstraction restricts actions available to one of the players. 
%If the player can win the abstract game, then it can also win the original 
%concrete game; however the opposite is not generally true.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Figure~\ref{fig:agt} shows an example abstract game
tree for the controller (abstract game trees for the environment are similar).  In the abstract game, 
the controller is required to pick actions
from the tree, starting from the root node.  After reaching a leaf, it
continues playing unrestricted.  The tree in Figure~\ref{fig:agt} restricts the
initial controller action to the set $\{a, d\}$.  After choosing action $a$,
the controller is required to play action $b$ in the next round.  It then reaches a leaf 
of the tree and continues playing unrestricted.
The root of the tree is annotated by the initial state $s$ of the abstract game and the 
bound $k$ on the number of rounds.
%Together with a state-round pair
%$\langle s,k\rangle$, an abstract game tree defines an \emph{abstract game}
%played from this state.
%Nodes of an abstract game tree are uniquely identified by the list of edge
%labels along the path from the root to the node.  
%We identify an abstract game tree with the set of its nodes.  
We denote $\textsc{nodes}(T)$ the set of all nodes of a tree $T$ and
$\textsc{leaves}(T)$ the subset of leaf nodes.

Given a controller (environment)
abstract game tree $T$ a \emph{partial strategy} $Strat: \textsc{nodes}(T) \rightarrow L_u$ 
($Strat: \textsc{nodes}(T) \rightarrow L_c$) labels each node of the tree with the opponent's
action to be played in this node.   
Figure~\ref{fig:strategy} shows an example partial strategy.  The environment 
starts by choosing action $\alpha$.  If the controller plays $a$, the
environment responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',k'\rangle=\textsc{outcome}(s, k, Strat, l)$, where $s'$ is the state obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf, and $i'$ is the number of rounds left in the game after reaching the leaf.
An opponent's partial strategy is winning if for all its outcomes $\langle s',k'\rangle$,
$s'$ is a winning state for the opponent in the concrete game with bound $k'$.

%The abstract game tree restricts moves available to the player, thus making the game easier
%to win for the opponent.  Hence, if the opponent does not have a winning partial strategy in
%the abstract game, it also does not have a winning strategy in the concrete game.

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, s, k, T, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$s, k, cand, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $k'$, $s'$, $\emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

Algorithm~\ref{alg:bounded} illustrates the bounded synthesis algorithm. The
algorithm takes a concrete game $G$ with bound $n$ as an implicit argument. In addition, it
takes a player $p$ (controller or environment), state $s$, bound
$k$ and an abstract game tree $T$ and returns a winning partial strategy
for $opponent(p)$, if one exists.  The initial invocation of the algorithm takes the
initial state $\langle I, n\rangle$ and an empty abstract game tree
$\emptyset$.  Initially the solver is playing on behalf of the environment since
that player takes the first move in every step.  The empty game tree does not
constrain opponent moves, hence solving such an abstraction is equivalent to
solving the original concrete game.  The algorithm is organised as a
counterexample-guided abstraction refinement (CEGAR) loop.  The first step of
the algorithm uses the \textsc{findCandidate} function, described below, to come 
up with a candidate partial strategy for $T$. If it fails to
find a strategy, this means that no winning partial strategy exists for $T$.
If, on the other hand, a candidate partial strategy is found, we need to verify
if it is indeed winning for $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.  
%{Figure~\ref{fig:spoiling}} shows a spoiling
%strategy discovered one of the leaves of the partial strategy.
